---
title: '[Reproducible Research](https://class.coursera.org/repdata-011), JHU via Coursera
  - Peer Assessment 1'
author: '[dpoet](https://github.com/dpoet/)'
date: "2015-FEB-15"
output:
  html_document:
    keep_md: yes
    theme: united
---

## [Reproducible Research](https://class.coursera.org/repdata-011), JHU via Coursera
### Peer Assessment 1    
## 2015-FEB-15  
## Author : [dpoet](https://github.com/dpoet/)  

 
 -----
  
## Introduction


  The assignment asks us to write a report that answers the questions below:  
  **Q1**: Loading and preprocessing the data  
  **Q2**: What is mean total number of steps taken per day?  
  **Q3**: What is the average daily activity pattern?  
  **Q4**: Imputing missing values  
  **Q5**: Are there differences in activity patterns between weekdays and weekends?  
  
  A series of tasks need to be performed in order to answer each question. I organized this work in four parts that answer all of the questions :
  
  -  In Part 1, "Loading and preprocessing the data", I answer the first question, so we are going to read the original data file (activity.zip), available from the following link: https://github.com/dpoet/RepData_PeerAssessment1/blob/master/activity.zip.  
  
     Also in Part 1, we will process and add columns that will be useful in further steps of the planned analysis.  
  
    
  -  In Part 2, "Exploring the data", I answer the second and the third questions, so we ignore missing data (which will be processed later) and start to understand our data by plotting it and calculating some useful measures by aggregating data by different dimensions (date, interval, etc.).  
  
  -  In Part 3, "Imputing missing values", I answer the fourth question, so we get back to the missing data: in this step, we take a look on the dataset's missing data and then devise a strategy for filling in all the missing values and derive a new dataset of complete cases (no missing data).  
  
  -  Finally, in Part 4, "Activity patterns on weekdays and weekends", I answer the fifth (and last) question, so we are interested in visualizing the activity patterns on weekdays and on weekends separately in order to compare them.
  
****  

### Part 1: Loading and preprocessing the data  


All data is loaded by function *LoadData()*, defined below.  
The *LoadData()* function already converts the variables data types as follows:  

  -    `steps`:   numeric  
  -    `date`:   Date    -   Format yyyy-mm-dd  
  - `interval`:   Factor  
  
Also, *LoadData()* will add the two variables below to the resulting data frame.  
These new variables will be useful later, during the analysis:  

  - `dayOfWeek`   :   Factor - The day of the week corresponding to the   `date` value.  
  - `weekendOrNot`:   Factor - A variable that indicates wether the corresponding   `date` value is on a weekend or on a weekday.  
  
  
  
  
  
```{r Question1a, echo = TRUE}
LoadData <- function(){ 
    # Constants (prefixed with a 'k')  
    ## Directories 
    kWORKING_DIR    <- getwd()
    kDATA_DIR       <- paste(kWORKING_DIR, "data"        , sep = "/")
    
    ## Data files
    kINPUT_FILE     <- paste(kWORKING_DIR, "activity.zip", sep = "/")
    kRAW_DATA_FILE  <- paste(kDATA_DIR,    "activity.csv", sep = "/")
    
    ## Data types
    kCOL_CLASSES    <- c("numeric", "Date", "numeric")
    kMISSING        <- "NA"
    
    ## Weekend and weekdays
    kWEEKEND_DAYS   <- c("Saturday", "Sunday")
    kIS_WEEKEND     <- "weekend"
    kNOT_WEEKEND    <- "weekday"
    
    ## Error messages
    kERROR_MESSAGES = list(
        errInputFileNotFound = "Fatal Error: the input file was not found.",
        errRawDataFileNotFound = "Fatal Error: the raw data file was not found."
    )
        
    # Variables
    hasRawData  <- FALSE
    varNames    <- NULL
    result      <- NULL
    
    
    if (! file.exists(kINPUT_FILE)){
        stop(kERROR_MESSAGES$errInputFileNotFound)
    } else {
        if (! file.exists(kRAW_DATA_FILE) ) {
            unzip(zipfile = kINPUT_FILE, exdir = kDATA_DIR)            
        }
        hasRawData <- file.exists(kRAW_DATA_FILE)
        
        if (! hasRawData){
            stop(kERROR_MESSAGES$errRawDataFileNotFound)
        } else {
            result <- read.csv(file             = kRAW_DATA_FILE, 
                               header           = TRUE,
                               na.strings       = kMISSING, 
                               colClasses       = kCOL_CLASSES
            )
            if (! is.null(result) ) {
                file.remove(kRAW_DATA_FILE) # Discards temporary "raw" file.
                varNames <- names(result)
                
                # Adds derived variables, useful during the analysis
                result$dayOfWeek        <- as.factor(weekdays(result[ , varNames[2] ]))
                result$weekendOrNot     <- as.factor(
                                            ifelse (result$dayOfWeek %in% kWEEKEND_DAYS, kIS_WEEKEND, kNOT_WEEKEND)
                                            )
   
            }
        }
    }
    result
}
```

All data is loaded into the `myData` data frame:
```{r Question1b, echo = TRUE}
    myData <- LoadData()
    str(myData)
    head(myData)
    tail(myData)
```
  
****  

### Part 2: Exploring the data  
  
#### Number of steps taken per day 

As stated in the Project instructions page, one can ignore the missing values in the dataset for this part of the assignment. 

So here we will consider only the observations that do NOT contain missing (`NA`) values:  
```{r CompleteSubset, echo = TRUE}
    myCompleteData <- myData[complete.cases(myData), ]  
```
  
  
  
  
##### The total number of steps taken each day  
The **total** number of steps taken per day was calculated and stored in the `totalStepsDay` variable.
```{r Question2.1, echo = TRUE, fig.path = "./figure/"}
    totalStepsDay        <- aggregate(x = myCompleteData$steps, by = list(myCompleteData$date), FUN = "sum")
    names(totalStepsDay) <- c("Day", "TotalSteps")
    hist(totalStepsDay$TotalSteps, col = "wheat3", 
         main = "Total Number of Steps Taken Each Day\n(Missing Data Ignored)", xlab = "Number of Steps")    
```
  
##### Mean and Median
The **mean** total number of steps taken per day was calculated and stored in the `meanStepsDay` variable.  
The **median** total number of steps taken per day was calculated and stored in the `medianStepsDay` variable.  
```{r Question2.2, echo = TRUE, fig.path = "./figure/", fig.width = 10, fig.height = 7}
    meanStepsDay    <- mean(totalStepsDay$TotalSteps)
    medianStepsDay  <- median(totalStepsDay$TotalSteps)
```
Thus:  

  -  The **mean** total number of steps taken per day is ~`r as.character(round(meanStepsDay,2))` ;  
  -  The **median** total number of steps taken per day is `r as.character(medianStepsDay)`.  

  
  
#### The average daily activity pattern
We calculate the mean values ( **across all days**) in order to obtain the average number of steps per 5-minute interval, which gives us the time series plot below:  
```{r Question3.1, echo = TRUE, fig.path="./figure/", fig.width=10}
    meanStepsInterval       <- aggregate(x = myCompleteData$steps, by = list(myCompleteData$interval), FUN = "mean")
    names(meanStepsInterval) <- c("Interval", "MeanSteps")       

    library(ggplot2) # Requires ggplot2 library in order to use qplot() function.
    qplot(data = meanStepsInterval, x = Interval, y = MeanSteps, 
          ylab = "Avg. Number of Steps per Interval", xlab = "5-minute Intervals", 
          main = "Average Daily Activity Pattern") + geom_line() + geom_smooth(method = "loess") + theme_light(base_size = 14) + theme(legend.position = "none")
```

##### Maximum Number of Steps
As we can see above, on average in the dataset, there is one 5-minute interval that contains the maximum number of steps:
```{r Question3.2, echo = TRUE}
    maxMeanSteps <- max(meanStepsInterval$MeanSteps)
    maxMeanStepsInterval <- with(
                                meanStepsInterval, 
                                meanStepsInterval[MeanSteps == maxMeanSteps , ]
                            )$Interval
```
The **maximum** average number of steps is observed in interval `r maxMeanStepsInterval` ( `maxMeanStepsInterval`), with a value of ~`r round(maxMeanSteps, 2)` ( `maxMeanSteps`).    
  
****

### 3. Imputing missing values  

##### The total number of missing values in the dataset can be calculated this way:  
```{r Question4.1, echo = TRUE}
    nrMissingValues <- nrow(myData[is.na(myData$steps) , ])
```
So, the total number of missing values in the dataset is `r  nrMissingValues`  (which
represents `r round(100 * nrMissingValues / nrow(myData), 2)`% of the observations).  
  
  
  
  
###### A strategy for filling in all of the missing values in the dataset.  
I chose to use the average number of steps by day of week and by interval:  

  - Step 1  
    Based on the dataset with no missing data ( `myCompleteData`): calculate and store the mean values for each pair of weekday and 5-minute interval;  
  - Step 2  
  For each missing value: given the day of the week and interval for which the measurement is missing, we look up the corresponding value and assign the result to the `steps` variable.  

**Step 1**: we first calculate the average values and store them in the   `meanWeekdayInterval` variable:  
```{r Question4.2a, echo = TRUE, fig.path="./figure/", fig.width=10, fig.height=10}
    meanWeekdayInterval <- aggregate(myCompleteData$steps, 
                                     by = list(myCompleteData$dayOfWeek, myCompleteData$interval),
                                     FUN ="mean")
    names(meanWeekdayInterval) <- c("wday", "itv", "meanSteps")
    head(meanWeekdayInterval, 7)
```

We can also visualize `meanWeekdayInterval` as follows:
```{r Question4.2b, echo = TRUE, fig.path="./figure/", fig.width=10, fig.height=10}
    qplot( data = meanWeekdayInterval, x = itv, y = meanSteps, facets = wday ~., col = wday, 
           main = "Avg. Steps by Week Day and 5-minute Interval", 
           ylab = "Average Steps", xlab = "5-minute Interval" ) + geom_line(lwd = 2) + theme_linedraw(base_size = 12) + theme(legend.position = "none")
```

**Step 2** : now we need to create a copy of the original dataset. That's the dataset in which missing values will be filled in.  


##### A *new dataset* that is equal to the original dataset but with the missing data filled in.  

```{r Question4.3, echo = TRUE}
    myNewData <- myData
    idxNA <- which(x = is.na(myNewData$steps), arr.ind = TRUE) # Obtains the index of NA values.
    for(i in idxNA)
    {
        weekDay <- as.character(myNewData[i, "dayOfWeek"])
        interv  <- myNewData[i, "interval"]
        
        # Use "weekDay" and "interv" to look up values in "meanWeekdayInterval" and
        # assign a new value to the "steps" variable.
        myNewData[i, "steps"] <- meanWeekdayInterval[meanWeekdayInterval$wday == weekDay & meanWeekdayInterval$itv == interv, "meanSteps"]  
    }
    
    head(myData, 20)        # The first 20 observations from the original dataset.
    head(myNewData, 20)     # The first 20 observations from the NEW dataset.
    tail(myData, 20)        # The last 20 observations from the original dataset.
    tail(myNewData, 20)     # The last 20 observations from the NEW dataset.
```
 
##### A histogram of the total number of steps taken each day 
```{r Question4.4a, echo = TRUE, fig.path="./figure/"}
    newTotalStepsDay        <- aggregate(x = myNewData$steps, by = list(myNewData$date), FUN = "sum")
    names(newTotalStepsDay) <- c("Day", "TotalSteps")
    hist(newTotalStepsDay$TotalSteps, col = "salmon3", main = "Total Number of Steps Taken Each Day", xlab = "Number of Steps")
```

##### Mean and Median
The **mean** total number of steps taken per day was calculated and stored in the `newMeanStepsDay` variable.  
The **median** total number of steps taken per day was calculated and stored in the `newMedianStepsDay` variable.  
```{r Question4.4b, echo = TRUE, fig.path = "./figure/", fig.width = 10, fig.height = 7}
    newMeanStepsDay    <- mean(newTotalStepsDay$TotalSteps)
    newMedianStepsDay  <- median(newTotalStepsDay$TotalSteps)
```
Thus:  

  -  The **mean** total number of steps taken per day is ~`r as.character(round(newMeanStepsDay,2))` (a little higher than the previous mean of ~`r as.character(round(meanStepsDay,2))`) ;  
  -  The **median** total number of steps taken per day is `r as.character(newMedianStepsDay)` (higher than the previous median of `r as.character(medianStepsDay)`).
  
As we can see, the use of more observations (`r nrow(newTotalStepsDay)` days instead of `r nrow(totalStepsDay)`) resulted in increased values for the median and the mean. Also, the median is not as close to the mean as before.

****





#### 4. Activity patterns on weekdays and weekends  
As mentioned in Part 1, the function *LoadData()*, used to load the dataset, also creates a new variable called `weekendOrNot`. This variable indicates wheter the corresponding day is a weekend (Saturday or Sunday) or a weekday. This variable enables us to visualize weekend and weekday behaviour separately.

##### Average number of steps on *weekends* and *weekdays*

```{r Question5.2, echo = TRUE, fig.path="./figure/", fig.width=10, fig.height=8}
    meanStepsWEWD          <- aggregate(x = myNewData$steps, 
                                        by = list(myNewData$interval, myNewData$weekendOrNot), 
                                        FUN = "mean")
    names(meanStepsWEWD)   <- c("Interval", "WeekendOrNot", "meanSteps")    
    
    qplot(data = meanStepsWEWD, x = Interval, y = meanSteps, 
          facets = WeekendOrNot ~ ., col = WeekendOrNot,
          xlab = "5-minute Interval", ylab = "Avg. Number of Steps") + geom_line(lwd = 2) + theme_linedraw(base_size = 20) + theme(legend.position = "none") + geom_smooth(method = "loess")
```

It is interesting to notice that:  
- On weekends, there is much more activity (greater number of steps) in the afternoon intervals than on week days ;  
- On week days, the "50-step" mark is reached just after interval 500, whereas on weekends, that happens closer to interval 1000.  
  
-----
